---
title: "Introducing specCurvieR"
author: "Zayne Sember"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
```

# What is specCurvieR?

specCurvieR is a work-in-progress R package aimed at making specification curve analysis easy, fast, and pretty.

# How do I install it?

devtools installation will be available soon. The goal is to have the package available on CRAN by the end of 2023.

# Why did you make it?

Data visualization and tinkering in R have been some of the most enjoyable parts of my research as a PhD student. The seeds for this package were planted when I took a course on replication in social science with Professor Gareth Nellis. An assignment involved performing a specification curve analysis for which Gareth provided us with some code to generate a specification curve. Not feeling like modifying someone else's code to get the final plot looking how I wanted, I was disappointed to see the available packages weren't much better. My biggest gripe was their use of base `R` plots which aren't exactly the prettiest ducklings and for which customization is a confusing hassle. Give me my `ggplot`! I want to arrange all the grobs! Long story short I gave in and used Gareth's code. Fast forward a couple years and I need to run a specification curve analysis but I can't find the assignment with Gareth's code--guess I'll just write it myself.

# Why should I use it over other packages?

specCurvieR seeks to provide everything alternative specification curve analysis packages offer with major improvements. Some features that set the package apart currently:
* Support for parallel computing to speed up model estimation
* Support for fixed effects
* Support for clustered and robust standard errors
* Option of a progress bar to monitor model estimation
* All plots are generated using ggplot2, allowing a high degree of customization
* Ability to compare various model fit parameters

# Can I see it in action?

I thought you'd never ask, let's walk through some examples.

The main function of the package is `sca()` (short for specification curve analysis, not the music). This is where you specify your model types, variables, controls, desired output, and whatnot.
```{r}
# Load packages
library(tidyverse)

# Load the specCurvieR
source("functions.R")

# Load some example data--CalCOFI's bottle database
# (https://calcofi.org/data/oceanographic-data/bottle-database/)
# and grab 500 observations from it
bottles <- sample_n(readRDS("Data/bottles.rds"), 500)

# Plenty of variables to play around with
names(bottles)
```
```{r}
# Suppose we're modelling the effect of salinity on ocean temperatures and
# want to understand how including the concentration of other chemicals affects 
# the model
s <- sca(y="T_degC", x="Salnty", controls=c("O2Sat", "ChlorA", "NH3uM",
                                            "NO2uM", "SiO3uM"), data=bottles)


# The function returns a dataframe containing a row for every possible
# combination of controls with all the information needed to generate plots.
s[1:4,1:4]
```

```{r}
# Lots of goodies in there, including the expression used to generate each
# model and indicator variables for the presence of each of the
# controls in that row's model
names(s)
```
```{r}
# Let's try that again but tell it produce some plots
s <- sca(y="T_degC", x="Salnty", controls=c("O2Sat", "ChlorA", "NH3uM",
                                            "NO2uM", "SiO3uM"), data=bottles,
         plot=T, plotSE="bar")

# Looks like our coefficient estimate for salinity changes quite a bit
# depending on what controls we include (a progress bar is displayed by default
# but does not show up when knitting)
s
```
```{r}
# Maybe we want things a little less colorful, let's set the control
# indicator ticks to all be black
s <- sca(y="T_degC", x="Salnty", controls=c("O2Sat", "ChlorA", "NH3uM",
                                            "NO2uM", "SiO3uM"), data=bottles,
         plot=T, colorControls=F)

s
```

